\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-24 k-planar.tex
Small fixes/clarifications in the proof of 1_planar_is_4_framed

\note{DW}{I suggest we use the language of ``embedded graph'' here}
For any integer $d\ge 3$, a (multi)graph $G$ is \defin{$d$-framed} if it has an embedding in $\R^2$ having a biconnected plane spanning (multi)subgraph $G_0$, whose facial cycles each have length at most $d$,  and such that the interior of each edge in $E(G)\setminus E(G_0)$ is contained in the interior of some face of $G_0$. The graph $G_0$ is called the \defin{frame} of $G$. This definition (for simple graphs) was introduced by \citet{BDGGMR}.

Every $d$-framed multigraph can be described by a $(2,\binom{d}{2})$-shortcut system applied to a planar multigraph (by adding a vertex inside each face $F$ of $G_0$ adjacent to the vertices of $F$). Thus \cref{ShortcutProduct} is applicable. However, for framed graphs we prove the following stronger result, where the treewidth bound on $H$ is best possible.\footnote{A multigraph $G$ is \defin{contained} in a graph $G'$ if the simple graph underlying $G$ is isomorphic to a subgraph of $G'$.}

@@

	The proof is by induction on the number of inner faces of $N'$.  A particular easy case occurs when $V(N')=V(F)$, for which we can simply take
	\note{DW}{Suggested revision, ``In the base case, $N'$ has only one inner face, and we can simply take ...} \note{PM}{I debated what to put here since, even when $N'$ has only one face, the rules below work for picking $\tau$.  The \emph{real} base case occurs when $N'$ has zero faces, but that's not something that ever happens since we only ever apply induction on the inner faces of $M$ (except $\tau$).  I settled on this ``base case'' since it's clear that, in this case, the partition $\mathcal{P}$ is already decided.}
  \note{PM}{The longwinded version is that induction is on the pair $(x,y)$ where $x=|V(N-F)|$ and $y$ is the number of inner faces of $N'$.  The base cases are $(0,y)$ for any $y$.}
	$\PP:=\{P_1,\ldots,P_k\}$ and verify that the preconditions of the lemma (the `setup') ensure that $\PP$ satisfies the requirements (i)--(iii):
	\begin{compactenum}[(i)]
		\item By definition, each of $P_1,\ldots,P_k$ is a part of $\mathcal{P}$.
		\item Assumption 4(b) ensures, for each $x\in V(H)$, the existence of $X_x$ and $Y_x$ such that $|X_x|\le d-3$ and $|Y_x\cap L_j|\le 3$ for each integer $j\ge 0$.
		\item The trivial tree decomposition of $H$ that has a single bag containing $\{x_1,\ldots,x_k\}$ is $3$-simple and, if $k=3$, has exactly one bag that contains $\{x_1,\ldots,x_3\}$.
	\end{compactenum}

\note{DW}{Now assume that $N'$ has at least two internal faces.}

In the general case, our $H$-partition $\mathcal{P}$ will include $P_1$, $P_2$ , $P_3$ and a (possibly empty) part $S$ that will be determined by three vertices $v_1v_2v_3$ that belong to a common inner face $\tau$ of $N'$.  We first explain how $v_1$, $v_2$ and $v_3$ are chosen and then explain how these are used to define $S$.  There is one case to consider for each possible value of $k$ (see \cref{boring_figure}):

@@

To understand the structure of $G$, it is helpful to consider the plane graph $G_0$ obtained by replacing each pair of edges $vw$ and $xy$ that cross at $p$ with four edges $vp$, $wp$, $xp$, and $yp$ meeting at a newly added \defin{dummy vertex} $p$. Let $F$ be a face of $G_0$ and let $W:=w_0,\ldots,w_{r-1}$ be the facial walk around $F$. In the following paragraphs, subscripts on the vertices in $W$ are implicitly taken modulo $r$. By edge maximality, if $W$ contains only non-dummy vertices, then it contains exactly three vertices and $F$ is bounded by three edges of $G$.

If $W$ contains a dummy vertex $w_i$, then neither $w_{i-1}$ nor $w_{i+1}$ is a dummy vertex since if $w_{i-1}$ (respectively $w_{i+1}$) were a dummy vertex then the edge of $G$ that contains $w_{i-1}w_i$ (respectively $w_iw_{i+1}$) would be involved in at least two crossings.  Therefore each dummy vertex $w_i$ of $W$ is surrounded by two non-dummy vertices $w_{i-1},w_{i+1}\in V(G)$.  Since $w_i$ has degree $4$ in $G_0$ and $F$ does not have two parallel edges on its boundary, $w_{i-1}\neq w_{i+1}$.  Since $G$ is edge-maximal the edge $w_{i-1}w_i$ is an edge of $G$ on the boundary of $F$.  Therefore, if $F$ has a dummy vertex $w_i$ in its facial walk $W$, then $W=w_{i-1},w_i,w_{i+1}$ and $F$ is bounded by three edges of $G_0$, each of which is contained in a different edge of $G$.

@@

Consider some pair of edges $vw$ and $xy$ that cross at point $p$, which a degree-4 vertex of $G_0$.  There are four faces $F_1,\ldots,F_4$ of $G_0$ that contain $p$ and, from the preceding discussion each of $vx$, $xw$, $wy$, and $yv$ is an uncrossed edge of $G$.  Therefore, by removing each pair of crossing edges from $G$ we obtain a plane multigraph $G'$ each of whose faces is bounded by three or four edges; see \cref{one_planar_example}(b).  Therefore $G$ is a $4$-framed graph with frame $G'$.
% Since, for any pair of crossing edges $\{vw,xy\}$ that we remove this way, $vxwy$ is a cycle in $G'$, $\dist_{G'}(v,w)\le 2$ and $\dist_{G'}(x,y)\le 2$.
% It follows immediately that, for any pair of vertices $a,b\in V(G)$, $\dist_{G'}(a,b)\le 2\dist_G(a,b)$.
\end{proof}

@@

As mentioned in \cref{Introduction}, \citet{DEJGMM21} prove results about adjacency labelling schemes for planar graphs. Stated in terms of universal graphs, their main theorem is interpreted as follows:

@@

\section{Orphaned Text}

\subsection{Text following \cref{gkPlanarProduct}}

Prior to this work, the strongest structural description of $k$-planar or $(g,k)$-planar graphs (or any of the other classes presented in \cref{examples}) was in terms of layered treewidth, which we now define.  A \defin{layered tree-decomposition} $(\mathcal{L},\mathcal{T})$ consists of a layering $\mathcal{L}$ and a tree-decomposition $\mathcal{T}$ of $G$. The layered width of $(\mathcal{L},\mathcal{T})$ is $\max\{|L\cap B|: L\in \mathcal{L},\, B\in \mathcal{T}\}$.  The \defin{layered treewidth} of $G$ is the minimum layered width of any layered tree-decomposition of $G$. \citet{dujmovic.morin.ea:layered} proved that planar graphs have layered treewidth at most 3, that graphs of Euler genus $g$ have layered treewidth at most $2g+3$, and more generally that a minor-closed class has bounded layered treewidth if and only if it excludes some apex graph. \citet{dujmovic.eppstein.ea:structure} show that every $k$-planar graph has layered treewidth at most $6(k+1)$, and more generally that every $(g,k)$-planar graph has layered treewidth at most $(4g+6)(k+1)$. It follows from this result that $(g,k)$-planar graphs have treewidth $O(\sqrt{(g+1)(k+1)n})$ and thus have balanced separators of the same order, which can also be concluded from the work of \citet{FP08}. In related work, \citet{grigoriev.bodlaender:algorithms} used structural results to obtain approximation algorithms for $(g,k)$-planar graphs, and \citet{PachToth97} determined the maximum number of edges in a $k$-planar graph (up to a constant factor).

If a graph class admits bounded layered partitions, then it also has bounded layered treewidth. In particular, if $\PP=(P_x:x\in V(H))$ is an $H$-partition of $G$ of layered width $\ell$ with respect to some layering $\mathcal{L}$ of $G$ and $(B_x:x\in V(T))$ is a width-$t$ tree-decomposition of $H$, then setting $C_x = \bigcup_{y\in B_x} P_y$ for each $x\in V(T)$ gives a tree-decomposition $(C_x:x\in V(T))$ of $G$ that has layered treewidth $(t+1)\ell$ \cite{DJMMUW20}. Therefore, any property that holds for graphs of bounded layered treewidth also holds for $G$. What sets layered partitions apart from layered treewidth is that they lead to constant upper bounds on the queue-number  and non-repetitive chromatic number, whereas for both these parameters, the best known upper bound obtainable via layered treewidth is $O(\log n)$; see \cref{Applications}.

%%%%%%%%%%%%%%%%%%%
\label{sec-k-planar}

%\note{DW}{I am pretty sure one can get treewidth 3 for $(g,1)$-planar graphs, combining \cref{d_framed_product_stucture} with the method in \citep{DHHW}. I think the best place for this result is reference \citep{DHHW} since the present paper pre-dates \citep{DHHW} (and this addition will increase the significance of \citep{DHHW}). Okay?}
%\note{PM}{Okay to this and to all similar questions that come later (which I've now commented out.)}
%\note{DW}{Robert is working on it.}

This section describes several examples of graph classes that can be obtained from a shortcut system typically applied to graphs of bounded Euler genus.

\subsection{Orphaned Theorems}

Theorems~\ref{PlanarProduct}(b),  \ref{ShortcutProduct} and  \ref{GenusProduct}(b) and \cref{MapShortcut,qn,p-centered,non-repetitive} imply the following result.

%\begin{thm}
%\label{PlaneMapPartition}
%\note{DW}{For every integer $d\geq 4$, }
%Every $d$-map graph $G$:
%	\begin{compactitem}
%		\item is contained in $H \boxtimes P \boxtimes K_{21d(d-3)}$ for some path $P$ and for some graph $H$ with $\tw(H)\leq 9$,
%		\item has queue-number $\qn(G) < 32225\, d(d-3)$.
%		\item has $p$-centered chromatic number $\chi_p(G) \leq 21d(d-3) (p+1)  \binom{p+9}{9}$,
%		\item has non-repetitive chromatic number $ \pi(G) \leq 21 \cdot 4^{10} d(d-3)$.
%	\end{compactitem}
%\end{thm}

\begin{thm}
\label{MapPartition}
For integers $g\geq 0$ and $d\geq 4$, if $\ell:=  7d(d-3)\, \max\{2g,3\}$ then every $(g,d)$-map graph $G$:
\begin{compactitem}
\item is contained in $H \boxtimes P \boxtimes K_{\ell}$ for some path $P$ and for some graph $H$ with $\tw(H)\leq 9$,
\item has queue-number $\qn(G) <  1535\, \ell $.
\item has $p$-centered chromatic number $\chi_p(G) \leq \ell (p+1)\,  \binom{p+9}{9}$,
\item has non-repetitive chromatic number $ \pi(G) \leq 4^{10}\,\ell $.
\end{compactitem}
\end{thm}

These results give the first constant upper bound on the non-repetitive chromatic number of map graphs, the first polynomial bounds on the $p$-centered chromatic number of map graphs, and the best known bounds on the queue-number of map graphs.

% \note{DW}{I am pretty sure one can get treewidth 3 here using \cref{framed} and the method in \citep{DHHW}. I think the best place for this result is reference \citep{DHHW} since the present paper pre-dates \citep{DHHW} (and this addition will increase the significance of \citep{DHHW}). Okay?}

% \subsection{Orphaned Section}
% 
% \cref{nearest-neighbour,kPlanarProduct,qn,p-centered} imply:
% 
% \begin{cor}
% \label{k-nn}
% For every integer $k\geq 1$ there exists integers $t\leq O(k^6)$ and $\ell\leq O(k^4)$ such that every $k$-nearest-neighbour graph:
% \begin{compactitem}
% \item is contained in $H\boxtimes P \boxtimes K_\ell$ for some graph $H$ with treewidth $t$ and some path $P$,
% \item has queue-number at most $2^{O(k^6)}$, and
% \item has $p$-centered chromatic number at most $\ell (p+1)\binom{p+t}{t}$.
% \end{compactitem}
% \end{cor}
% 
% \cref{non-repetitive,k-nn} also give bounds on the non-repetitive chromatic number of a $k$-nearest neighbour graph $G$. However, the bound is weak, since $G$ has maximum degree at most $6k$, implying that $\pi(G) \leq (36+o(1))k^2$ by a result of \citet{DJKW16}.

\subsection{Generalisations}
\label{Generalisations}

% As mentioned above, product structure theorems have been established for several minor-closed classes in addition to planar graphs. The first generalises \cref{PlanarProduct} for graphs of bounded Euler genus.
%
% \begin{thm}[\citep{DJMMUW20,UWY,DHHW}]
% \label{GenusProduct}
% Every graph of Euler genus $g$ is contained in:
% \begin{compactenum}[(a)]
% \item $H  \boxtimes P$ for some graph $H$ of treewidth at most $2g+6$  and some path $P$.
% \item $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some graph $H$ of treewidth at most $3$ and for some path $P$.
% \end{compactenum}
% \end{thm}

\citet{DJMMUW20} generalised \cref{GenusProduct} for apex-minor-free graphs as follows.

\begin{thm}[\citep{DJMMUW20}]
\label{ApexMinorFree}
For every apex graph $X$, there exists $c\in\mathbb{N}$ such that every $X$-minor-free graph is contained in $H\boxtimes P$ for some graph $H$ with $\tw(H)\leq c$ and some path $P$.
\end{thm}

The assumption that $X$ is apex is needed in \cref{ApexMinorFree}, since if the class of $X$-minor-free graphs has a product structure theorem analogous to \cref{PlanarProduct}, then $X$ is apex \citep{DJMMUW20}. On the other hand, \citet{DEMWW22} proved a product structure theorem for bounded degree graphs in any minor-closed class.

\begin{thm}[\citep{DEMWW22}]
	\label{MinorFreeDegree}
	For every graph $X$ there exists $c\in\mathbb{N}$ such that for every $\Delta\in\mathbb{N}$, every $X$-minor-free graph $G$ with maximum degree at most $\Delta$ is contained in $H\boxtimes P$ for some graph $H$ with $\tw(H) \leq c\Delta$ and for some path $P$.
\end{thm}

%The next three results are immediate corollaries of \cref{GenusProduct,ApexMinorFree,ShortcutProduct,MinorFreeDegree}.
%
%\begin{thm}
%Let $\SS$ be a $(k,d)$-shortcut system for a graph $G$ of Euler genus $g$. Then $G^\SS$ is contained in $H\boxtimes P$ for some graph $H$ of treewidth at most $d(k^3+3k)\binom{k+2g+8}{2g+8}-1$ and for some path $P$.
%\end{thm}
%
%\begin{thm}
%For every apex graph $X$ and for all integers $k,d\geq 1$, there is an integer $c$ such that for every $X$-minor-free graph $G$ and for every $(k,d)$-shortcut system $\SS$ for $G$, $G^\SS\subseteq H\boxtimes P$ for some graph $H$ with $\tw(H)\leq c$ and for some path $P$.
%\end{thm}
%
%\begin{thm}
%For every graph $X$ and for all integers $k,\Delta\geq 1$, there is an integer $c$ such that for every $X$-minor-free graph $G$ with maximum degree $\Delta$ and for every $(k,\Delta)$-shortcut system $\SS$ for $G$, $G^\SS \subseteq H \boxtimes P$ for some graph $H$ with $\tw(H)\leq c$ and for some path $P$.
%\end{thm}
%
%\note{DW}{delete the above three theorems, they are never used}

%\subsection*{Note Added in Proof} Subsequent to the initial release of this paper, the treewidth 8 bound in \cref{PlanarProduct} was improved to 6 by \citet{UWY}, which implies that the treewidth $2g+8$ bound in \cref{GenusProduct}(a) can be improved to $2g+6$. Simmilarly, the treewidth 4 bound in \cref{GenusProduct}(b) was improved to 3 by \citet{DHHW}. \note{DW}{Say more about the consequence for the bounds in the rest of the paper}

>>>>

% \note{DW}{I suggest we use the language of ``embedded graph'' here}
For any integer $d\ge 3$, an embedded (multi)graph $G$ is \defin{$d$-framed} if it has a biconnected plane spanning (multi)subgraph $G_0$, whose facial cycles each have length at most $d$,  and such that the interior of each edge in $E(G)\setminus E(G_0)$ is contained in the interior of some face of $G_0$. The embedded (multi)graph $G_0$ is called the \defin{frame} of $G$. This definition (for simple graphs) was introduced by \citet{BDGGMR}.

Every $d$-framed multigraph can be described by a $(2,d(d-3)/2)$-shortcut system applied to a plane multigraph by adding a vertex inside each face $F$ of $G_0$ adjacent to the vertices of $F$ and creating a shortcut between each non-consecutive pair of vertices of $F$. Thus \cref{ShortcutProduct} is applicable. However, for framed graphs we prove the following stronger result, where the treewidth bound on $H$ is best possible.\footnote{A multigraph $G$ is \defin{contained} in a graph $G'$ if the simple graph underlying $G$ is isomorphic to a subgraph of $G'$.}

@@

	The proof is by induction on the number of inner faces of $N'$.  A particular easy case occurs $N'$ has no inner vertices, i.e., $V(N')=V(F)$. In this case we can simply take
	\note{DW}{Suggested revision, ``In the base case, $N'$ has only one inner face, and we can simply take ...} 
  % \note{PM}{I debated what to put here since, even when $N'$ has only one face, the rules below work for picking $\tau$.  The \emph{real} base case occurs when $N'$ has zero faces, but that's not something that ever happens since we only ever apply induction on the inner faces of $M$ (except $\tau$).  I settled on this ``base case'' since it's clear that, in this case, the partition $\mathcal{P}$ is already decided.}
  \note{PM}{The longwinded version is that induction is on $(x,y)$ where $x$ is the number of inner vertices of $N'$ and $y$ is the number of inner faces of $N'$.  The base cases are $(0,y)$ for any $y$.  Unfortunately, we can't just induct on $x$ because it doesn't always decrease.}
	$\PP:=\{P_1,\ldots,P_k\}$ and verify that the preconditions of the lemma (the `setup') ensure that $\PP$ satisfies the requirements (i)--(iii):
	\begin{compactenum}[(i)]
		\item By definition, each of $P_1,\ldots,P_k$ is a part of $\mathcal{P}$.
		\item Assumption 4(b) ensures, for each $x\in V(H)$, the existence of $X_x$ and $Y_x$ such that $|X_x|\le d-3$ and $|Y_x\cap L_j|\le 3$ for each integer $j\ge 0$.
		\item The trivial tree decomposition of $H$ that has a single bag containing $\{x_1,\ldots,x_k\}$ is $3$-simple and, if $k=3$, has exactly one bag that contains $\{x_1,\ldots,x_3\}$.
	\end{compactenum}

% \note{DW}{Now assume that $N'$ has at least two internal faces.}

We now move onto the case in which $N'$ contains at least one inner vertex.
Our $H$-partition $\mathcal{P}$ will include $P_1$, $P_2$ , $P_3$ and a (possibly empty) part $S$ that will be determined by three vertices $v_1v_2v_3$ that belong to a common inner face $\tau$ of $N'$.  We first explain how $v_1$, $v_2$ and $v_3$ are chosen and then explain how these are used to define $S$.  There is one case to consider for each possible value of $k$ (see \cref{boring_figure}):

@@

Refer to \cref{one_planar_example}.
To understand the structure of $G$, it is helpful to consider the plane graph $G_0$ obtained by replacing each pair of edges $vw$ and $xy$ that cross at $p$ with four edges $vp$, $wp$, $xp$, and $yp$ meeting at a newly added \defin{dummy vertex} $p$. Let $F$ be a face of $G_0$ and let $W:=w_0,\ldots,w_{r-1}$ be the facial walk around $F$. In the following paragraphs, subscripts on the vertices in $W$ are implicitly taken modulo $r$. By edge maximality, if $W$ contains only non-dummy vertices, then it contains exactly three vertices and $F$ is bounded by three edges of $G$.

If $W$ contains a dummy vertex $w_i$, then neither $w_{i-1}$ nor $w_{i+1}$ is a dummy vertex since if $w_{i-1}$ (respectively $w_{i+1}$) were a dummy vertex then the edge of $G$ that contains $w_{i-1}w_i$ (respectively $w_iw_{i+1}$) would be involved in at least two crossings.  Therefore each dummy vertex $w_i$ of $W$ is flanked by two non-dummy vertices $w_{i-1},w_{i+1}\in V(G)$.  Since $w_i$ has degree $4>1$ in $G_0$ and no two edges incident to a common vertex cross each other, $w_{i-1}\neq w_{i+1}$.  Since $G$ is edge-maximal the edge $w_{i-1}w_{i+1}$ is an edge of $G$ on the boundary of $F$.  Therefore, if $F$ has a dummy vertex $w_i$ in its facial walk $W$, then $W=w_{i-1},w_i,w_{i+1}$ and $F$ is bounded by three edges of $G_0$, each of which is contained in a different edge of $G$.

@@

Consider some pair of edges $vw$ and $xy$ that cross at point $p$, which is a degree-4 vertex of $G_0$.  Since no pair of edges incident to a common vertex cross each other, $v$, $w$, $x$, and $y$ are all distinct.  There are four faces $F_1,\ldots,F_4$ of $G_0$ with $p$ on their boundary and, from the preceding discussion each of $vx$, $xw$, $wy$, and $yv$ is an uncrossed edge of $G$.  Therefore, by removing each pair of crossing edges from $G$ we obtain a plane multigraph $G'$ each of whose faces is bounded by three or four edges.  Therefore $G$ is a $4$-framed graph with frame $G'$.
% Since, for any pair of crossing edges $\{vw,xy\}$ that we remove this way, $vxwy$ is a cycle in $G'$, $\dist_{G'}(v,w)\le 2$ and $\dist_{G'}(x,y)\le 2$.
% It follows immediately that, for any pair of vertices $a,b\in V(G)$, $\dist_{G'}(a,b)\le 2\dist_G(a,b)$.
\end{proof}

@@

As mentioned in \cref{Introduction}, \citet{DEJGMM21,EJM} prove results about adjacency labelling schemes for planar graphs. Stated in terms of universal graphs, their main theorem is interpreted as follows:

@@

% \section{Orphaned Text}
% 
% \subsection{Text following \cref{gkPlanarProduct}}
% 
% Prior to this work, the strongest structural description of $k$-planar or $(g,k)$-planar graphs (or any of the other classes presented in \cref{examples}) was in terms of layered treewidth, which we now define.  A \defin{layered tree-decomposition} $(\mathcal{L},\mathcal{T})$ consists of a layering $\mathcal{L}$ and a tree-decomposition $\mathcal{T}$ of $G$. The layered width of $(\mathcal{L},\mathcal{T})$ is $\max\{|L\cap B|: L\in \mathcal{L},\, B\in \mathcal{T}\}$.  The \defin{layered treewidth} of $G$ is the minimum layered width of any layered tree-decomposition of $G$. \citet{dujmovic.morin.ea:layered} proved that planar graphs have layered treewidth at most 3, that graphs of Euler genus $g$ have layered treewidth at most $2g+3$, and more generally that a minor-closed class has bounded layered treewidth if and only if it excludes some apex graph. \citet{dujmovic.eppstein.ea:structure} show that every $k$-planar graph has layered treewidth at most $6(k+1)$, and more generally that every $(g,k)$-planar graph has layered treewidth at most $(4g+6)(k+1)$. It follows from this result that $(g,k)$-planar graphs have treewidth $O(\sqrt{(g+1)(k+1)n})$ and thus have balanced separators of the same order, which can also be concluded from the work of \citet{FP08}. In related work, \citet{grigoriev.bodlaender:algorithms} used structural results to obtain approximation algorithms for $(g,k)$-planar graphs, and \citet{PachToth97} determined the maximum number of edges in a $k$-planar graph (up to a constant factor).
% 
% If a graph class admits bounded layered partitions, then it also has bounded layered treewidth. In particular, if $\PP=(P_x:x\in V(H))$ is an $H$-partition of $G$ of layered width $\ell$ with respect to some layering $\mathcal{L}$ of $G$ and $(B_x:x\in V(T))$ is a width-$t$ tree-decomposition of $H$, then setting $C_x = \bigcup_{y\in B_x} P_y$ for each $x\in V(T)$ gives a tree-decomposition $(C_x:x\in V(T))$ of $G$ that has layered treewidth $(t+1)\ell$ \cite{DJMMUW20}. Therefore, any property that holds for graphs of bounded layered treewidth also holds for $G$. What sets layered partitions apart from layered treewidth is that they lead to constant upper bounds on the queue-number  and non-repetitive chromatic number, whereas for both these parameters, the best known upper bound obtainable via layered treewidth is $O(\log n)$; see \cref{Applications}.
% 
% 
% 
% %%%%%%%%%%%%%%%%%%%
% \label{sec-k-planar}
% 
% 
% %\note{DW}{I am pretty sure one can get treewidth 3 for $(g,1)$-planar graphs, combining \cref{d_framed_product_stucture} with the method in \citep{DHHW}. I think the best place for this result is reference \citep{DHHW} since the present paper pre-dates \citep{DHHW} (and this addition will increase the significance of \citep{DHHW}). Okay?}
% %\note{PM}{Okay to this and to all similar questions that come later (which I've now commented out.)}
% %\note{DW}{Robert is working on it.}
% 
% 
% 
% 
% 
% This section describes several examples of graph classes that can be obtained from a shortcut system typically applied to graphs of bounded Euler genus.
% 
% \subsection{Orphaned Theorems}
% 
% Theorems~\ref{PlanarProduct}(b),  \ref{ShortcutProduct} and  \ref{GenusProduct}(b) and \cref{MapShortcut,qn,p-centered,non-repetitive} imply the following result.
% 
% %\begin{thm}
% %\label{PlaneMapPartition}
% %\note{DW}{For every integer $d\geq 4$, }
% %Every $d$-map graph $G$:
% %	\begin{compactitem}
% %		\item is contained in $H \boxtimes P \boxtimes K_{21d(d-3)}$ for some path $P$ and for some graph $H$ with $\tw(H)\leq 9$,
% %		\item has queue-number $\qn(G) < 32225\, d(d-3)$.
% %		\item has $p$-centered chromatic number $\chi_p(G) \leq 21d(d-3) (p+1)  \binom{p+9}{9}$,
% %		\item has non-repetitive chromatic number $ \pi(G) \leq 21 \cdot 4^{10} d(d-3)$.
% %	\end{compactitem}
% %\end{thm}
% 
% \begin{thm}
% \label{MapPartition}
% For integers $g\geq 0$ and $d\geq 4$, if $\ell:=  7d(d-3)\, \max\{2g,3\}$ then every $(g,d)$-map graph $G$:
% \begin{compactitem}
% \item is contained in $H \boxtimes P \boxtimes K_{\ell}$ for some path $P$ and for some graph $H$ with $\tw(H)\leq 9$,
% \item has queue-number $\qn(G) <  1535\, \ell $.
% \item has $p$-centered chromatic number $\chi_p(G) \leq \ell (p+1)\,  \binom{p+9}{9}$,
% \item has non-repetitive chromatic number $ \pi(G) \leq 4^{10}\,\ell $.
% \end{compactitem}
% \end{thm}
% 
% These results give the first constant upper bound on the non-repetitive chromatic number of map graphs, the first polynomial bounds on the $p$-centered chromatic number of map graphs, and the best known bounds on the queue-number of map graphs.
% 
% % \note{DW}{I am pretty sure one can get treewidth 3 here using \cref{framed} and the method in \citep{DHHW}. I think the best place for this result is reference \citep{DHHW} since the present paper pre-dates \citep{DHHW} (and this addition will increase the significance of \citep{DHHW}). Okay?}
% 
% 
% % \subsection{Orphaned Section}
% % 
% % \cref{nearest-neighbour,kPlanarProduct,qn,p-centered} imply:
% % 
% % \begin{cor}
% % \label{k-nn}
% % For every integer $k\geq 1$ there exists integers $t\leq O(k^6)$ and $\ell\leq O(k^4)$ such that every $k$-nearest-neighbour graph:
% % \begin{compactitem}
% % \item is contained in $H\boxtimes P \boxtimes K_\ell$ for some graph $H$ with treewidth $t$ and some path $P$,
% % \item has queue-number at most $2^{O(k^6)}$, and
% % \item has $p$-centered chromatic number at most $\ell (p+1)\binom{p+t}{t}$.
% % \end{compactitem}
% % \end{cor}
% % 
% % \cref{non-repetitive,k-nn} also give bounds on the non-repetitive chromatic number of a $k$-nearest neighbour graph $G$. However, the bound is weak, since $G$ has maximum degree at most $6k$, implying that $\pi(G) \leq (36+o(1))k^2$ by a result of \citet{DJKW16}.
% 
% 
% 
% 
% \subsection{Generalisations}
% \label{Generalisations}
% 
% % As mentioned above, product structure theorems have been established for several minor-closed classes in addition to planar graphs. The first generalises \cref{PlanarProduct} for graphs of bounded Euler genus.
% %
% % \begin{thm}[\citep{DJMMUW20,UWY,DHHW}]
% % \label{GenusProduct}
% % Every graph of Euler genus $g$ is contained in:
% % \begin{compactenum}[(a)]
% % \item $H  \boxtimes P$ for some graph $H$ of treewidth at most $2g+6$  and some path $P$.
% % \item $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some graph $H$ of treewidth at most $3$ and for some path $P$.
% % \end{compactenum}
% % \end{thm}
% 
% \citet{DJMMUW20} generalised \cref{GenusProduct} for apex-minor-free graphs as follows.
% 
% \begin{thm}[\citep{DJMMUW20}]
% \label{ApexMinorFree}
% For every apex graph $X$, there exists $c\in\mathbb{N}$ such that every $X$-minor-free graph is contained in $H\boxtimes P$ for some graph $H$ with $\tw(H)\leq c$ and some path $P$.
% \end{thm}
% 
% The assumption that $X$ is apex is needed in \cref{ApexMinorFree}, since if the class of $X$-minor-free graphs has a product structure theorem analogous to \cref{PlanarProduct}, then $X$ is apex \citep{DJMMUW20}. On the other hand, \citet{DEMWW22} proved a product structure theorem for bounded degree graphs in any minor-closed class.
% 
% \begin{thm}[\citep{DEMWW22}]
% 	\label{MinorFreeDegree}
% 	For every graph $X$ there exists $c\in\mathbb{N}$ such that for every $\Delta\in\mathbb{N}$, every $X$-minor-free graph $G$ with maximum degree at most $\Delta$ is contained in $H\boxtimes P$ for some graph $H$ with $\tw(H) \leq c\Delta$ and for some path $P$.
% \end{thm}
% 
% %The next three results are immediate corollaries of \cref{GenusProduct,ApexMinorFree,ShortcutProduct,MinorFreeDegree}.
% %
% %\begin{thm}
% %Let $\SS$ be a $(k,d)$-shortcut system for a graph $G$ of Euler genus $g$. Then $G^\SS$ is contained in $H\boxtimes P$ for some graph $H$ of treewidth at most $d(k^3+3k)\binom{k+2g+8}{2g+8}-1$ and for some path $P$.
% %\end{thm}
% %
% %\begin{thm}
% %For every apex graph $X$ and for all integers $k,d\geq 1$, there is an integer $c$ such that for every $X$-minor-free graph $G$ and for every $(k,d)$-shortcut system $\SS$ for $G$, $G^\SS\subseteq H\boxtimes P$ for some graph $H$ with $\tw(H)\leq c$ and for some path $P$.
% %\end{thm}
% %
% %\begin{thm}
% %For every graph $X$ and for all integers $k,\Delta\geq 1$, there is an integer $c$ such that for every $X$-minor-free graph $G$ with maximum degree $\Delta$ and for every $(k,\Delta)$-shortcut system $\SS$ for $G$, $G^\SS \subseteq H \boxtimes P$ for some graph $H$ with $\tw(H)\leq c$ and for some path $P$.
% %\end{thm}
% %
% %\note{DW}{delete the above three theorems, they are never used}
% 
% 
% 
% 
% %\subsection*{Note Added in Proof} Subsequent to the initial release of this paper, the treewidth 8 bound in \cref{PlanarProduct} was improved to 6 by \citet{UWY}, which implies that the treewidth $2g+8$ bound in \cref{GenusProduct}(a) can be improved to $2g+6$. Simmilarly, the treewidth 4 bound in \cref{GenusProduct}(b) was improved to 3 by \citet{DHHW}. \note{DW}{Say more about the consequence for the bounds in the rest of the paper}
